Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-03-14 paper/paper.tex
an embarassing typo and extended the appendix on mnist-like
edits after independent proofreading

% general intro text with motivation
Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. With continual miniaturization ever more
applications can be found in embedded systems. Common embedded applications include
on-device image recognition and signal processing. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge due to storage, real-time throughput, and arithmetic (computational)
complexity restrictions \citep{he_amc:_2018}. Therefore compression methods for achieving
high model sparsity and numerical efficiency without losing much in performance are
especially relevant.
% Solving the storage and artihmetic constraints encourages development of model compression
% and sparsification methods, that focus on favourable trade-off between performance and size.

Inherently complex-valued nature of data in signal processing, specifically, acoustic or radio
frequency signal analysis, has been the key driver behind adoption of $\cplx$-networks.
%
\citet{hirose_complex-valued_2009} argues that the key merit of $\cplx$-valued neural
networks is simultaneous phase rotation and amplitude adjustment, intrinsic to multiplication
in $\cplx$ field. Despite inherently higher arithmetic complexity, which stems from up to four
floating point operations per single $\cplx$ multiplication \citep{monning_evaluation_2018}
$\cplx$-arithmetic reduces ineffective degrees of freedom in comparison with an analogous
double-dimensional $\real$-valued network. \citet{hirose_complex-valued_2009} list numerous
applications related to acoustic and radio signal processing and demonstrate superiority of
$\cplx$-networks in the task of shallow landmine detection using ground penetrating radar
imaging. More recently they were applied to music transcription and speech recognition
\citep{trabelsi_deep_2017}. \citet{arjovsky_unitary_2016} propose $\cplx$-valued unitary-RNNs
to deal with exploding~/~vanishing gradients and apply the recurrent models to long-term
dependence prediction tasks. \citet{wisdom_full-capacity_2016} expand on their work by optimizing
recurrent weights on the $\cplx^{n \times n}$ Stiefel manifold. And \citet{wolter_complex_2018}
investigate different $\cplx$-valued gating mechanisms unitary-RNNs and conduct experiments
on human motion prediction and music transcription. \citet{yang_complex_2019} develop complex
transformer, with $\cplx$-attention and complex encoder-decoder, and apply it to music
transcription and wireless signal classification.

@@

% segue into bayesian inference
Bayesian inference is a very rich field of research with a large body of research and
techniques that enable practical quantification of predictive, parameter, or model
uncertainty and its propagation into solutions of the problems of interest.
%
Dropout resembles averaging models with shared parameters: each model is weighted equally
since masks are randomly sampled, unlike Bayesian model averaging, which weighs according to
the posterior distribution. \citet{kingma_variational_2015} specifically consider dropout
as a Bayesian Inference technique: a principled framework of reasoning about uncertainty,
which updates prior beliefs in accordance with the likelihood of empirical observations
into posterior belief. Bayesian methods with suitable modelling and prior assumptions, such
as Variational Dropout, can be used towards model sparsification
\citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.

@@

The posterior distribution itself is intractable, save for the relatively simple cases, and
therefore exact inference is traded for tractability and scalability offered by Variational
Inference. Instead of deriving $p(\omega \mid D)$ using the rule, this approach recasts
posterior inference into variational optimization problem: finding $q$ that is close to
$p(\omega \mid D)$ in terms of a proximity score $\rho$
\begin{equation}  \label{eq:variational-progam}
  q_*
    \in \arg \min_{\theta} \rho\bigl(
      q_\theta(\omega), p(\omega \mid D)
    \bigr)
    \,,
\end{equation}
over a tractable parametric family of distributions on the parameter space $
  \mathcal{Q} = \{Q_\theta(d\omega) \colon \theta \in \Theta\}
$, $\theta$ -- generic variational parameter. The class $\mathcal{Q}$ is picked so that
its members have tractable densities $
  q_\theta(\omega) d\omega
$, can be sampled from and possess tractable $\log$-derivatives $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$ \citep{williams_simple_1992} or can be represented as differentiable transformations
of non-parametric noise \citep{kingma_auto-encoding_2014,figurnov_implicit_2019}.
% (reparameterization trick, pathwise gradieent) $q_{\theta}(d\omega)$ is a push-forward
% of $p(d\varepsilon)$ by a differentiable map $g(\varepsilon; \theta)$.
% % http://stillbreeze.github.io/REINFORCE-vs-Reparameterization-trick/#fn:1
%
The natural choice for $\rho$ is Kullback-Leibler divergence of density $q$ from $p$,
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
the main reason being that it satisfies the following identity for well-behaved densities:
% $(q, p, \pi)$
\begin{multline}  \label{eq:kl-div-master}
  % \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
  \log p(D)
    - KL(q \| p(\omega \mid D))
    \\ = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q \| \pi)
    \,.
\end{multline}
This identity yields an equivalent problem for \eqref{eq:variational-progam}:
maximizing the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(D \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where the prior and likelihood are allowed to depend on explicit parameters. The most
commonly used algorithm to optimize \eqref{eq:elbo_general} is EM-algorithm. Other
variational objectives are possible, provided $p(\omega \mid D)$ is probed only through
$\log p(D \mid \omega)$ and $\log \pi(\omega)$ \citep{ranganath_operator_2018}.

@@

With a special family of posterior approximation, variational inference can be used as
a regularization method and as a model sparsification technique. \citet{kingma_variational_2015}
consider Dropout \citep{hinton_improving_2012}, DropConnect \citep{wan_regularization_2013},
and Gaussian dropout \citep{srivastava_dropout_2014,wang_fast_2013} through the lens
of Bayesian inference methods and proposes \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\xi
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \xi \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
%
Variational Dropout assumes fully factorized Gaussian approximate posterior $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. The divergence term in \eqref{eq:elbo} unravels into a sum of the following
analytically intractable terms:
\begin{equation}  \label{eq:improper-kl-div-real}
  KL(q_\theta \| \pi)
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximate this expression by a non-linear polynomial
regression for truncated $\alpha \in (0, 1)$. \citet{molchanov_variational_2017}
refine the approximation by using a nonlinear regression based on sigmoid and soft-plus.
% see p.6 sec 3.4 of kingma_variational_2015
We independently verify their approximation against Monte-Carlo estimates for $\alpha$
varying over a fine logarithmic grid and the exact expression for the derivative of
\eqref{eq:improper-kl-div-real}, which we derive in appendix~\ref{sub:real-chisq-grad}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
%
% (in) it is in poor taste to brag about this begin the first time anyone has derived a
% derivative of an intractable function.

Being the ratio of the squared mean to the effective variance, $\alpha_j$ regulates the
relevance of the parameter $\omega_j$ it is associated to. Thus \citet{molchanov_variational_2017}
focus on the model sparsification capabilities of Variational Dropout, optimizing $\alpha$
for each individual parameter. Flexibility in specification of $q_\theta$ enables structured
dropout with shared $\alpha$ across groups of parameters $\omega$. Another Bayesian dropout
method, called \textit{Automatic Relevance Determination}, is developed in \citep{kharitonov_variational_2018}.
The key difference is that the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper
Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This method, known as \textit{Empirical Bayes},
fits a prior distribution while performing variational inference. Maximizing \eqref{eq:elbo}
over $\tau$, holding other parameters fixed, yields $
  \tau^* = {(\mu^2 + \sigma^2)}^{-1}
$, whence
\begin{equation}  \label{eq:ard-kl-div-real}
  KL(q_\theta \| \pi)
    = \log{\bigl(1 + \tfrac1{\alpha} \bigr)}
    \,.
\end{equation}

@@

In this section we briefly review $\cplx$-valued networks, their distinction from the
ordinary $\real$-valued networks.

\citet{trabelsi_deep_2017} outline the building blocks for deep $\cplx$-valued networks
and describe suitable representation and operations including convolutional and dense layers,
$\cplx$-valued activations, complex batch-normalization and suitable weight initialization.
Their approach makes a complex-valued network into an intricately connected $\real$-valued
computational graph that respects $\cplx$-arithmetic.
%
In particular, natural identification of $\cplx$ and $\real^2$ implies that linear layers $
  L \colon \cplx^m \to \cplx^n
$ act upon their inputs as follows:
\begin{equation}  \label{eq:cplx-lin-op}
  {(\real^n)}^{2}
    \to {(\real^m)}^{2}
    \colon (u, v)
      \mapsto \Bigl(
        P u - Q v,
        P v + Q u
      \Bigr)
    \,,
\end{equation}
where $
  P, Q \colon \real^{n} \to \real^{m}
$ are unique operators such that $L = P + j Q$. A $\cplx$-convolutional layer with kernel
$W$ can be implemented as two $\real$-convolutions with kernels $\Re{W}$ and $\Im{W}$ using
\eqref{eq:cplx-lin-op}. Activations include trigonometric functions and representation-dependent
non-linearities \citep{hirose_complex-valued_2009}: polar (phase-amplitude) $
  z \mapsto \sigma(\lvert z \rvert) e^{j \arg z}
$ or planar (real-imaginary) $
  z \mapsto \sigma(\Re z) + j \sigma(\Im z)
$ for some $\real$-valued non-linearity $\sigma$, possibly with learnable parameters
\citep{trabelsi_deep_2017,wolter_complex_2018}.

Complex networks are, in general, non-holomorphic, i.e. not $\cplx$-differentiable, which is
exacerbated by the loss being real-valued. Indeed a $\cplx$-differentiable real-valued function is
necessarily trivial by Cauchy-Riemann conditions. This issue is dealt with by employing Wirtinger,
or $\cplx\real$ calculus, which generalizes holomorphic calculus to non-holomorphic functions
of $\cplx$-argument \citep{adali_complex-valued_2011,boeddeker_computation_2019}. It satisfies
product and chain rules, respects complex conjugation and linearity, but most importantly the
$\cplx\real$-differential of a function on $\cplx$ coincides with its classical differential as
a function on $\real^2$. Essentially, it allows straightforward retrofiting of $\cplx$-valued
networks into existing $\real$ deep learning auto-differentiation frameworks. It was considered
as a basis for $\cplx$ version of back-propagation by \citet{benvenuto_complex_1992} and
\citet{trabelsi_deep_2017}, (see appendix~\ref{sub:wirtinger_calculus} for details).
% \todo{read on holomorphic nets}

@@

For $\cplx$-variational dropout we propose using mean-field proper $\cplx$-Gaussian approximate
posterior, e.g. parameters $W \in \cplx^{n\times m}$ of a dense linear layer follow
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W \sim q(W)
    % = \prod_{ij} q(W_{ij})
    = \prod_{ij} \mathcal{N}^{\cplx}(
      W_{ij} \mid
        \mu_{ij},
        \alpha_{ij} \lvert \mu_{ij} \rvert^2,
        % \sigma^2_{ij},
        0
    )
  \,,
\end{equation}
and we use similar factorized approximation for convolutions and other, effectively, linear
layers. For this approximation the relation coefficients \eqref{eq:cplx-gauss-trick} are
zero. We use MAP estimates of bias terms, i.e. assume an independent fully factorized Dirac
approximation.

@@

Consider the following parameterization of $W$: $
  W = \mu \odot \xi
$ where $
  \xi_{ij} \sim \mathcal{N}(1, \alpha_{ij})
$, yet $\mu \in \cplx^{n \times m}$. This case corresponds to inference regarding
relevance coefficients $\xi$ rather than the parameters themselves after. Each weight
under this parameterization is an independent degenerate $\cplx$-Gaussian distribution
\begin{equation}  \label{eq:real-scal-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij},
      \, \alpha_{ij} \lvert \mu_{ij} \rvert^2,
      \, \alpha_{ij} \mu_{ij}^2
    \bigr)
  \,,
\end{equation}
which implies by \eqref{eq:cplx-gauss-trick} the following:
\begin{equation}  \label{eq:real-scal-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}\bigl(
      % e_i^\top \mu x + b,
      b_i + \sum_{j=1}^m \mu_{ij} x_j,
      \, \sum_j \alpha_{ij} \lvert x_{ij} \mu_{ij}\rvert^2,
      \, \sum_j \alpha_{ij} (x_{ij} \mu_{ij})^2
    \bigr)
    \,.
\end{equation}
The KL-divergence term for this approximation coincides with \eqref{eq:improper-kl-div-real},
which is approximated by \citet{molchanov_variational_2017}. The major drawback of this method
is inapplicability of the additive noise reparameterization in \eqref{eq:real-scal-gauss-trick},
which disentangles the gradient with respect to $\mu$ from the local noise.

@@

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.

@@

The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. 
%
The benefit of the ``fine-tune'' stage appears to be strongly influenced by the compression
rate. In this experiment and more prominently in MusicNet (sec.~\ref{sub:musicnet}), we
can observe that the more ``holes'' Bayesian dropout poked in the model, the more regularized
it becomes, the less capacity it maintains for overfitting and the more ``fine-tune'' serves
as extended training.

>>>>

% general intro text with motivation
Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. With continual miniaturization ever more
applications can be found in embedded systems. Common embedded applications include
on-device image recognition and signal processing. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge due to storage, real-time throughput, and arithmetic (computational)
complexity restrictions \citep{he_amc:_2018}. Therefore, compression methods for achieving
high model sparsity and numerical efficiency without losing much in performance are
especially relevant.
% Solving the storage and artihmetic constraints encourages development of model compression
% and sparsification methods, that focus on favourable trade-off between performance and size.

Inherently complex-valued nature of data in signal processing, specifically, acoustic or radio
frequency signal analysis, has been the key driver behind the adoption of $\cplx$-networks.
%
\citet{hirose_complex-valued_2009} argues that the key merit of $\cplx$-valued neural
networks is simultaneous phase rotation and amplitude adjustment, intrinsic to multiplication
in $\cplx$ field. Despite inherently higher arithmetic complexity, which stems from up to four
floating point operations per single $\cplx$ multiplication \citep{monning_evaluation_2018}
$\cplx$-arithmetic reduces ineffective degrees of freedom in comparison with an analogous
double-dimensional $\real$-valued network. \citet{hirose_complex-valued_2009} lists numerous
applications related to acoustic and radio signal processing and demonstrates superiority of
$\cplx$-networks in the task of shallow landmine detection using ground penetrating radar
imaging. More recently they have been applied to music transcription and speech recognition
\citep{trabelsi_deep_2017}. \citet{arjovsky_unitary_2016} propose $\cplx$-valued unitary-RNNs
to deal with exploding~/~vanishing gradients and apply the recurrent models to long-term
dependence prediction tasks. \citet{wisdom_full-capacity_2016} expand on their work by optimizing
recurrent weights on the $\cplx^{n \times n}$ Stiefel manifold. And \citet{wolter_complex_2018}
investigate different $\cplx$-valued gating mechanisms for unitary-RNNs and conduct experiments
on human motion prediction and music transcription. \citet{yang_complex_2019} develop complex
transformer, with $\cplx$-attention and complex encoder-decoder, and apply it to music
transcription and wireless signal classification.

@@

% segue into bayesian inference
Bayesian inference is a very rich field with a large body of research and techniques that
enable practical quantification of predictive, parameter, or model uncertainty and its 
propagation into solutions of the problems of interest.
%
Dropout resembles averaging models with shared parameters: each model is weighted equally
since masks are randomly sampled, unlike Bayesian model averaging, which weighs according to
the posterior distribution. \citet{kingma_variational_2015} specifically consider dropout
as a Bayesian Inference technique: a principled framework of reasoning about uncertainty,
which updates prior beliefs in accordance with the likelihood of empirical observations
into posterior belief. Bayesian methods with suitable modelling and prior assumptions, such
as Variational Dropout, can be used towards model sparsification
\citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.

@@

The posterior distribution itself is intractable, save for the relatively simple cases, and
therefore exact inference is traded for tractability and scalability offered by Variational
Inference. Instead of deriving $p(\omega \mid D)$ using the Bayes rule, this approach recasts
posterior inference into variational optimization problem: finding $q$ that is close to
$p(\omega \mid D)$ in terms of a proximity score $\rho$
\begin{equation}  \label{eq:variational-progam}
  q_*
    \in \arg \min_{\theta} \rho\bigl(
      q_\theta(\omega), p(\omega \mid D)
    \bigr)
    \,,
\end{equation}
over a tractable parametric family of distributions on the parameter space $
  \mathcal{Q} = \{Q_\theta(d\omega) \colon \theta \in \Theta\}
$, $\theta$ -- generic variational parameter. The class $\mathcal{Q}$ is picked so that
its members have tractable densities $
  q_\theta(\omega) d\omega
$, can be sampled from and have either tractable $\log$-derivatives $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$ \citep{williams_simple_1992} or can be represented as differentiable transformations
of non-parametric noise \citep{kingma_auto-encoding_2014,figurnov_implicit_2019}.
% (reparameterization trick, pathwise gradieent) $q_{\theta}(d\omega)$ is a push-forward
% of $p(d\varepsilon)$ by a differentiable map $g(\varepsilon; \theta)$.
% % http://stillbreeze.github.io/REINFORCE-vs-Reparameterization-trick/#fn:1
%
The natural choice for $\rho$ is Kullback-Leibler divergence of density $q$ from $p$,
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
the main reason being that it satisfies the following identity for well-behaved densities:
% $(q, p, \pi)$
\begin{multline}  \label{eq:kl-div-master}
  % \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
  \log p(D)
    - KL(q \| p(\omega \mid D))
    \\ = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q \| \pi)
    \,.
\end{multline}
This identity yields an equivalent problem for \eqref{eq:variational-progam}:
maximizing the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(D \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where the prior and likelihood are allowed to depend on explicit parameters. The most
commonly used algorithm to optimize \eqref{eq:elbo_general} is EM-algorithm. Other
variational objectives are possible, provided $p(\omega \mid D)$ is probed only through
$\log p(D \mid \omega)$ and $\log \pi(\omega)$ \citep{ranganath_operator_2018}.

@@

With a special family of posterior approximation, variational inference can be used as
a regularization method and as a model sparsification technique. \citet{kingma_variational_2015}
consider Dropout \citep{hinton_improving_2012}, DropConnect \citep{wan_regularization_2013},
and Gaussian dropout \citep{srivastava_dropout_2014,wang_fast_2013} through the lens
of Bayesian inference methods and propose \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\xi
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \xi \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
%
Variational Dropout assumes fully factorized Gaussian approximate posterior $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. The divergence term in \eqref{eq:elbo} unravels into a sum of the following
analytically intractable terms:
\begin{equation}  \label{eq:improper-kl-div-real}
  KL(q_\theta \| \pi)
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximate this expression by a non-linear polynomial
regression over $\alpha \in (0, 1)$. \citet{molchanov_variational_2017} refine the
approximation by using a nonlinear regression based on sigmoid and soft-plus.
% see p.6 sec 3.4 of kingma_variational_2015
We independently verify their approximation against Monte-Carlo estimates for $\alpha$
varying over a fine logarithmic grid and the exact expression for the derivative of
\eqref{eq:improper-kl-div-real}, which we derive in appendix~\ref{sub:real-chisq-grad}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
%
% (in) it is in poor taste to brag about this begin the first time anyone has derived a
% derivative of an intractable function.

Being the ratio of the squared mean to the effective variance, $\alpha_j$ regulates the
relevance of the parameter $\omega_j$ it is associated to. Thus, \citet{molchanov_variational_2017}
focus on the model sparsification capabilities of Variational Dropout, optimizing $\alpha$
for each individual parameter. Flexibility in specification of $q_\theta$ enables structured
dropout with shared $\alpha$ across groups of parameters $\omega$. Another Bayesian dropout
method, called \textit{Automatic Relevance Determination}, has been developed in \citep{kharitonov_variational_2018}.
The key difference is that the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper
ARD Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This method, known as \textit{Empirical Bayes},
fits a prior distribution while performing variational inference. Maximizing \eqref{eq:elbo}
over $\tau$, holding other parameters fixed, yields $
  \tau^* = {(\mu^2 + \sigma^2)}^{-1}
$, whence
\begin{equation}  \label{eq:ard-kl-div-real}
  KL(q_\theta \| \pi)
    = \log{\bigl(1 + \tfrac1{\alpha} \bigr)}
    \,.
\end{equation}

@@

In this section we briefly review $\cplx$-valued networks, and their distinction from the
ordinary $\real$-valued networks.

\citet{trabelsi_deep_2017} outline the building blocks for deep $\cplx$-valued networks
and describe suitable representation and operations including convolutional and dense layers,
$\cplx$-valued activations, complex batch-normalization and suitable weight initialization.
Their approach transforms a complex-valued network into an intricately connected $\real$-valued
computational graph that respects $\cplx$-arithmetic.
%
In particular, natural identification of $\cplx$ and $\real^2$ implies that linear layers $
  L \colon \cplx^m \to \cplx^n
$ act upon their inputs as follows:
\begin{equation}  \label{eq:cplx-lin-op}
  {(\real^n)}^{2}
    \to {(\real^m)}^{2}
    \colon (u, v)
      \mapsto \Bigl(
        P u - Q v,
        P v + Q u
      \Bigr)
    \,,
\end{equation}
where $
  P, Q \colon \real^{n} \to \real^{m}
$ are unique operators such that $L = P + j Q$. A $\cplx$-convolutional layer with kernel
$W$ can be implemented as two $\real$-convolutions with kernels $\Re{W}$ and $\Im{W}$ using
\eqref{eq:cplx-lin-op}. Activations include trigonometric functions and representation-dependent
non-linearities \citep{hirose_complex-valued_2009}: polar (phase-amplitude) $
  z \mapsto \sigma(\lvert z \rvert) e^{j \arg z}
$ or planar (real-imaginary) $
  z \mapsto \sigma(\Re z) + j \sigma(\Im z)
$ for some $\real$-valued non-linearity $\sigma$, possibly with learnable parameters
\citep{trabelsi_deep_2017,wolter_complex_2018}.

Complex networks are, in general, non-holomorphic, i.e. not $\cplx$-differentiable, which is
exacerbated by the loss being real-valued. Indeed a $\cplx$-differentiable real-valued function is
necessarily trivial by Cauchy-Riemann conditions. This issue is dealt with by employing Wirtinger,
or $\cplx\real$ calculus, which generalizes holomorphic calculus to non-holomorphic functions
of $\cplx$-argument \citep{adali_complex-valued_2011,boeddeker_computation_2019}. It satisfies
the product and chain rules, respects complex conjugation and linearity, but most importantly the
$\cplx\real$-differential of a function on $\cplx$ coincides with its classical differential as
a function on $\real^2$. Essentially, it allows straightforward retrofiting of $\cplx$-valued
networks into existing $\real$ deep learning auto-differentiation frameworks. It was considered
as a basis for $\cplx$ version of back-propagation by \citet{benvenuto_complex_1992} and
\citet{trabelsi_deep_2017}, (see appendix~\ref{sub:wirtinger_calculus} for details).
% \todo{read on holomorphic nets}

@@

For $\cplx$-variational dropout we propose using mean-field proper $\cplx$-Gaussian approximate
posterior and MAP estimates of bias terms, i.e. parameters $W \in \cplx^{n\times m}$ of a dense
linear layer follow
% We use MAP estimates of bias terms, i.e. assume an independent fully factorized Dirac approximation.
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W \sim q(W)
    % = \prod_{ij} q(W_{ij})
    = \prod_{ij} \mathcal{N}^{\cplx}(
      W_{ij} \mid
        \mu_{ij},
        \alpha_{ij} \lvert \mu_{ij} \rvert^2,
        % \sigma^2_{ij},
        0
    )
  \,,
\end{equation}
and we use similar factorized approximation for convolutions and other, effectively, linear
layers. For this approximation the relation coefficients \eqref{eq:cplx-gauss-trick} are
zero.

@@

Consider the following parameterization of $W$: $
  W = \mu \odot \xi
$ where $
  \xi_{ij} \sim \mathcal{N}(1, \alpha_{ij})
$, yet $\mu \in \cplx^{n \times m}$. This case corresponds to inference regarding
relevance coefficients $\xi$ rather than the parameters themselves after. Each weight
under this parameterization is an independent degenerate $\cplx$-Gaussian distribution
\begin{equation}  \label{eq:real-scal-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij},
      \, \alpha_{ij} \lvert \mu_{ij} \rvert^2,
      \, \alpha_{ij} \mu_{ij}^2
    \bigr)
  \,,
\end{equation}
which implies by \eqref{eq:cplx-gauss-trick} the following:
\begin{equation}  \label{eq:real-scal-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}\bigl(
      % e_i^\top \mu x + b,
      b_i + \sum_{j=1}^m \mu_{ij} x_j,
      \, \sum_j \alpha_{ij} \lvert x_{ij} \mu_{ij}\rvert^2,
      \, \sum_j \alpha_{ij} (x_{ij} \mu_{ij})^2
    \bigr)
    \,.
\end{equation}
The KL-divergence term for this approximation coincides with \eqref{eq:improper-kl-div-real},
which is approximated by \citet{molchanov_variational_2017}. The major drawback of this method
is that the additive noise reparameterization in \eqref{eq:real-scal-gauss-trick}, which
disentangles the gradient with respect to $\mu$ from the local noise, cannot be applied.

@@

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that have been studied in prior research on complex-valued networks.

@@

The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. 
%
The benefit of the ``fine-tune'' stage appears to be strongly influenced by the compression
rate. We can observe in this and, more prominently, in MusicNet experiment (sec.~\ref{sub:musicnet})
that higher compression rate render the model more regularized, leaving less capacity for
overfitting.
% more ``holes'' Bayesian dropout ``poked'' makes the ``fine-tune'' serve as extended training.
